REDO Simulating Dwarf Galaxy Evolution with Individual Star Feedback and Chemical Evolution I: Methods and Test Case

### 2017-11-21 main.tex
added a citation to Reissl et al. 2017, further updates in section 2

\subsubsection{Stellar Winds}
\label{sec:stellar winds}
Stellar winds are important sources of enrichment and feedback in galaxies at both early times from massive stars and late times from AGB stars. Although the energy injected via winds over the lifetime of a cluster of stars is much less than that of supernovae and radiation, stellar winds are potentially important sources of pre-supernova feedback. In both AGB phase winds and massive star winds we inject the feedback energy in a 3-cell radius spherical region centered on the emitting star particle assuming full thermalization of the wind kinetic energy. The total injected energy in timestep $\Delta t$ is therefore $E_{\rm wind} = \dot{M}\Delta t \times v^2_{\rm wind} + E_{\rm th}$, where $E_{\rm th}$ is the thermal energy of the ejected gas mass ($\dot{M}\Delta t$) given the star's interpolated $T_{\rm eff}$. We use a monte-carlo sampling method to inject mass and energy into the surrounding cells to account for partial cell overlap at the edge if the injection region that is a result of mapping a spherical injection region to cubic grid cells.

We assume constant mass loss rates for all winds as set by the yields tables and either the lifetime of the star (for massive stars) or the length of the AGB phase (for low mass stars). 
%Properly modeling stellar winds requires sub-parsec resolution, which is problematic in galaxy-scale simulations. 
Massive star stellar winds have typical velocities on order 10$^{3}$ km s$^{-1}$ \citep{Leitherer1992}. Satisfying the Courant time step becomes prohibitively expensive following this gas ($dt \sim$~100~yr), particularly as stellar wind bubbles have typical temperatures of 2 $\times$ 10$^{6}$ K \citep{Weaver1977}. For this reason, we adopt the common simplification of reducing the wind velocity \citep[e.g][]{Offner2015}. In our case, we fix massive star stellar winds to $v_{\rm wind} = 1$~km~s$^{-1}$ for stars above 8 M$_{\odot}$. Our initial tests show that turning off energy injection from stellar winds does not significantly affect the global star formation rate of our galaxies. Due to the substantial additional computational expense of following stellar winds for Gyr timescales, we reserve examining the detailed importance of winds to future work. These points are discussed in more detail in Section \ref{sec:stellar winds discussion}.

@@

Stars that undergo an AGB phase only deposit their feedback at the end of their lives, as determined by the PARSEC evolution tracks. AGB wind velocities vary dramatically over their relatively short lifetimes, but are typically on the order of 10 km s$^{-1}$. For simplicity, we adopt a fixed AGB phase wind velocity of 20 km s$^{-1}$ for all AGB stars.

\subsubsection{Core Collapse Supernovae}
\label{sec: core collapse}
Stars between  $M_{\rm thresh} = 8$ M$_{\odot}$ and 25 M$_{\odot}$ end their lives as core collapse supernovae, ejecting mass and yields as determined by the NuGrid stellar yield tables at 10$^{51}$ erg of thermal energy. Due to the high resolution (1.5 - 2.5 pc depending on galaxy size) of our simulations, we generally well resolve the Sedov phase of each supernova explosion (see Appendix~\ref{appendix:supernova}). We find injecting thermal energy alone in a 3 cell radius spherical region around the star particle to be sufficient to resolve the supernovae explosions. We continue to track any remaining stellar mass after the SNe as a massive ``remnant" tracer particle. In future work these particles can be used to self-consistently account for more exotic sources of feedback and chemical enrichment through X-ray binaries and neutron-star-neutron-star merger events which, while rare, could be important in long term galaxy evolution.
\textbf{Should I reference a few recent hydro sims that included these feedback sources? - AE}

\subsubsection{Type Ia Supernova}
\label{sec:Type Ia}
We continue to track low mass stars ($M < 8$ M$_{\odot}$) after their death as white dwarf (WD) particles, marking a subset as Type Ia supernova progenitors. At the end of their life, we assign a new mass to these particles following the initial to final mass relation of \citet{Salaris2009}. \textbf{I originally reproduced the equations here, but it is currently commented out. - AE}
%or, 
%\[
%M_{\rm WD} = 
%\begin{cases}
%0.134 M_{\rm i} + 0.331 & 1.7~{\rm M_{\odot}} \le M_{\rm *} \le 4~{\rm M_{\odot}} \\
%0.047 M_{\rm i} + 0.679 & 4~{\rm M_{\odot}} \le M_{\rm *}
%\end{cases}
%\]
%where $M_{\rm *}$ is the initial mass of the WD progenitor.
We follow the common assumption that progenitor stars with masses between 3 and 8 M$_{\odot}$ form WD's that are Type Ia progenitors \citep[see][ and references therein]{Cote2017}. We compute the probability that a given Type Ia progenitor will explode as a function of time using an observationally motivated delay time distribution (DTD) model. 
%\textbf{I went in detail on the DTD model here out of frustration of the lack of detail in other works, but I can trim this if needed (AE):} 
The SNIa rate is taken as a power law in time, $\Psi (t) \propto t^{-\beta}$, whose slope and normalization are observables, and $\eta$ the fraction of stars which can form SN Ia progenitors that \textit{will} explode as a SN Ia within a Hubble time. The value for $\eta$ depends on the normalization of the DTD (an observable), as well as one's choice of IMF and mass range of potential SN Ia progenitors (as given above). This fraction is given as
\begin{equation}
\eta = \frac{N_{\rm Ia}}{M_{\rm SF}} \frac{\int_{M_{\rm min}}^{M_{\rm max}} m (dN/dm) dm }{\int_{M_{1}}^{M_{2}} (dN/dm) dm},
\end{equation}
where $N_{\rm Ia}/M_{\rm SF}$ is the DTD normalization, which represents the fraction of stars that form SN Ia progenitors, $M_{\rm min}$ and $M_{\rm max}$ are the lower and upper bounds of the IMF ($dN/dm$), and $M_{1}$ and $M_{2}$ are the lower and upper bounds of the range of stars that can form Type Ia candidates. The DTD slope, $\beta$, is of order unity, with typical values between 1.0 and 1.2 (see \citet{Maoz2014} for a recent review). $N_{\rm Ia}/M_{\rm SF}$ represents the number of SN Ia per mass of star formation; this can be derived by taking observed values of the SNIa rate and integrating over a Hubble time. Typical values for this are on order of 10$^{-3}$ M$_{\odot}^{-1}$ \citep{Maoz2014}. We adopt 1.2 \citet{Maoz2010} and 0.8$\times 10^{-3}$ M$_{\odot}^{-1}$ \citet{GraurMaoz2013} for our fiducial values of $\beta$ and $N_{\rm Ia}/M_{\rm SF}$. Given our choice of IMF, and with $M_{\rm min} = 1$~M$_{\odot}$, $M_{\rm max}=100$~M$_{\odot}$, $M_{1}=3$~M$_{\odot}$, and $M_{2}=8$~M$_{\odot}$, this gives $\eta = 0.041$.

@@

\subsubsection{Ionizing Radiation}
\label{sec:ionizing radiation}
Radiation feedback (ionization, ionization heating, and radiation pressure) is an important source of pre-SN feedback in galaxies. HII regions carved out by stellar radiation change the ISM structure where SNe eventually explode, generally increasing their dynamical importance. However, when accounting for angular effects, radiation can also allow energy from SNe to dissipate more readily through channels carved through dense clouds. Radiation feedback effects have been included with various approximations in a wide range of simulations \citep[e.g.][]{OppenheimerDave2006, Krumholz2007, HopkinsQuataertMurray2012, Agertz2013, Renaud2013, Stinson2013, Roskar2014, Ceverino2014, FIRE, AgertzKravtsov2015, Forbes2016, Hu2016, Hu2017, FIRE2}, with limited application in full radiation hydrodynamics simulations \citep{WiseAbel2012,Wise2012a,Wise2014,Kim2013a, Kim2013b,Pawlik2013,Rosdahl2015,Aubert2015,Ocvirk2016,BaczynskiGloverKlessen2015,Pawlik2017} due the additional computational expense of direct ray tracing. As we seek a complete accounting of stellar feedback physics, we follow HI and HeI ionizing radiation from our stars through ray tracing methods described below. In this particular work, we use two different simulations of the same galaxy to demonstrate the importance of following radiation feedback with full radiative transfer methods (see Section~\ref{sec:IC}).

@@

We adopt the \ion{H}{1} and \ion{He}{1} ionizing photon rates for each star using the OSTAR2002 \citep{Lanz2003} grid of O-type stellar models, appropriate for $M_{*} \gtrsim 15$~M$_{\odot}$ at solar metallicity\footnote{The exact stellar mass range on the OSTAR2002 grid is model dependent and a function of metallicity}. We use linear interpolation in stellar effective temperature, surface gravity, and metallicity to compute the ionizing photon fluxes and rates for each star. Stars less massive than about 15 M$_{\odot}$ are generally off of the OSTAR2002 grid, and massive stars with sub-solar metallicity generally have effective temperatures above the OSTAR2002 grid. In this case, we integrate a black body spectrum at $T_{\rm eff}$ to obtain the ionizing photon fluxes, but normalize the result to be continuous with the OSTAR2002 grid (see Appendix \ref{appendix:radiation}). \citep{Dale2012} demonstrated that uncertainties on order of a few in the ionizing radiation rates are not significant in giant molecular cloud (GMC) scale simulations.
%Generally, the ionizing fluxes from high temperature black bodies underestimate the OSTAR2002 flux, and lower temperature black bodies overestimate the flux by a factor of 2-5 and 10-1000 respectively.

Instead of assigning a fixed ionizing photon energy across all sources, we integrate over each star's blackbody curve to compute the average ionizing photon energy individually for each source (see Appendix~\ref{appendix:radiation}). The average energy for \ion{H}{1} and \ion{He}{1} ionizing photons changes significantly over the OSTAR2002 temperature range ($log(T_{4,\rm{eff}} [K]) \in \left[2.75,5.5\right]$), ranging from 15.72 eV to 20.07 eV and 26.52 eV to 31.97 eV respectively.

We also include the effects of radiation pressure on HI. This has been shown to be important in suppressing the star formation rates of dwarf galaxies by influencing turbulence and the dense gas content of the ISM \citep{WiseAbel2012,Ceverino2014}. We ignore the absorption of ionizing radiation by dust and re-radiation in the infrared. This is included in other models \citep[e.g.][]{Rosdahl2015,FIRE,FIRE2} as this may increase by a factor of a few to several the effective radiation pressure \citep{ZhangDavis2017}. However, the importance of multiple scattering is still unclear. Other works have shown the effect to only increase the radiation pressure by a factor on order of unity \citep{Krumholz2012,Krumholz2013}. Due to these uncertainties, and given that we focus on low metallicity dwarf galaxies with low dust content, we ignore this effect.

>>>>

\subsubsection{Stellar Winds}
\label{sec:stellar winds}
Stellar winds are important sources of enrichment and feedback in galaxies at both early times from massive stars and late times from AGB stars. Although the energy injected via winds over the lifetime of a cluster of stars is much less than that from supernovae and radiation, stellar winds are potentially important sources of pre-supernova feedback. 
%mm For both AGB winds and massive stellar winds
    In models that include feedback energy from stellar winds from both AGB and 
    massive stars, 
we inject the feedback energy in a three-cell radius spherical region centered on the emitting star particle, assuming full thermalization of the wind kinetic energy. The total injected energy in timestep $\Delta t$ is therefore $E_{\rm wind} = \dot{M}\Delta t \times v^2_{\rm wind} + E_{\rm th}$, where $E_{\rm th}$ is the thermal energy of the ejected gas mass ($\dot{M}\Delta t$) given the star's interpolated $T_{\rm eff}$. We use a Monte Carlo sampling method to inject mass and energy into the surrounding cells to account for partial cell overlap at the edge of the injection region that is a result of mapping a spherical injection region to cubic grid cells.
We assume constant mass loss rates for all winds as set by the yields tables over either the lifetime of the star (for massive stars) or the length of the AGB phase (for low mass stars). 
%Properly modeling stellar winds requires sub-parsec resolution, which is problematic in galaxy-scale simulations.

Massive stellar winds have typical velocities of order 10$^{3}$ km s$^{-1}$ \citep{Leitherer1992}. Satisfying the Courant time step becomes prohibitively expensive following this gas ($dt \sim$~100~yr), particularly as stellar wind bubbles have typical temperatures of 2 $\times$ 10$^{6}$ K \citep{Weaver1977}. For this reason, we adopt the common simplification of reducing the wind velocity \citep[e.g][]{Offner2015}. In our case, we fix massive stellar winds to $v_{\rm wind} = 1$~km~s$^{-1}$ for stars above 8 M$_{\odot}$. Our initial tests show that turning off energy injection from stellar winds like this does not significantly affect the global star formation rate of our galaxies. Due to the substantial additional computational expense of following stellar winds for gigayear timescales, we reserve examining the detailed importance of winds to future work. These points are discussed in more detail in Section \ref{sec:stellar winds discussion}.

@@

Stars that only undergo an AGB phase deposit their feedback at the end of their lives, as determined by the PARSEC evolution tracks. AGB wind velocities vary dramatically over their relatively short lifetimes, but are typically on the order of 10 km s$^{-1}$. For simplicity, we adopt a fixed wind velocity of 20 km s$^{-1}$ for all AGB stars.

\subsubsection{Core Collapse Supernovae}
\label{sec: core collapse}
Stars between  $M_{\rm thresh} = 8$ M$_{\odot}$ and 25 M$_{\odot}$ end their lives as core collapse supernovae, ejecting mass and yields as determined by the NuGrid stellar yield tables at 10$^{51}$ erg of thermal energy. Due to the high resolution (1.5--2.5 pc depending on galaxy size) of our simulations, we generally resolve the Sedov phase of each supernova explosion well (see Appendix~\ref{appendix:supernova}). We find injecting thermal energy alone in a three-cell radius spherical region around the star particle to be sufficient to resolve the supernova explosions. We continue to track any remaining stellar mass after the supernovae as a massive ``remnant" tracer particle. In future work these particles can be used to self-consistently account for more exotic sources of feedback and chemical enrichment through X-ray binaries and neutron-star-neutron-star merger events which, while rare, could be important in long term galaxy evolution.
\textbf{Should I reference a few recent hydro sims that included these feedback sources? - AE  
%mm
Can't hurt if they are ready to hand. Not high priority, though. M-M}

\subsubsection{Type Ia Supernova}
\label{sec:Type Ia}
We continue to track low mass stars ($M < 8$ M$_{\odot}$) after their death as white dwarf (WD) particles, marking a subset as Type Ia supernova progenitors. At the end of their life, we assign a new mass to these particles following the initial-to-final-mass relation of \citet{Salaris2009}.
%mm \textbf{I originally reproduced the equations here, but it is currently commented out. - AE}
%or, 
%\[
%M_{\rm WD} = 
%\begin{cases}
%0.134 M_{\rm i} + 0.331 & 1.7~{\rm M_{\odot}} \le M_{\rm *} \le 4~{\rm M_{\odot}} \\
%0.047 M_{\rm i} + 0.679 & 4~{\rm M_{\odot}} \le M_{\rm *}
%\end{cases}
%\]
%where $M_{\rm *}$ is the initial mass of the WD progenitor.
We follow the common assumption that progenitor stars with masses between 3 and 8 M$_{\odot}$ form WD's that are Type Ia progenitors \citep[see][ and references therein]{Cote2017}.

We compute the probability that a given Type Ia progenitor will explode as a function of time using an observationally motivated delay time distribution (DTD) model. 
%\textbf{I went in detail on the DTD model here out of frustration of the lack of detail in other works, but I can trim this if needed (AE):} 
The SNIa rate is taken as a power law in time, $\Psi (t) \propto t^{-\beta}$, whose slope and normalization are observables, and $\eta$ the fraction of stars which can form SN Ia progenitors that \textit{will} explode as a SN Ia within a Hubble time. The value for $\eta$ depends on the normalization of the DTD (an observable), as well as one's choice of IMF and mass range of potential SN Ia progenitors (as given above). This fraction is given as
\begin{equation}
\eta = \frac{N_{\rm Ia}}{M_{\rm SF}} \frac{\int_{M_{\rm min}}^{M_{\rm max}} m (dN/dm) dm }{\int_{M_{1}}^{M_{2}} (dN/dm) dm},
\end{equation}
where $N_{\rm Ia}/M_{\rm SF}$ is the DTD normalization, which represents the fraction of stars that form SN Ia progenitors, $M_{\rm min}$ and $M_{\rm max}$ are the lower and upper bounds of the IMF ($dN/dm$), and $M_{1}$ and $M_{2}$ are the lower and upper bounds of the range of stars that can form Type Ia candidates. The DTD slope, $\beta$, is of order unity, with typical values between 1.0 and 1.2 (see \citet{Maoz2014} for a recent review). $N_{\rm Ia}/M_{\rm SF}$ represents the number of SN Ia per mass of star formation; this can be derived by taking observed values of the SNIa rate and integrating over a Hubble time. Typical values for this are on order of 10$^{-3}$ M$_{\odot}^{-1}$ \citep{Maoz2014}. We adopt 1.2 \citep{Maoz2010} and 0.8$\times 10^{-3}$ M$_{\odot}^{-1}$ \citep{GraurMaoz2013} for our fiducial values of $\beta$ and $N_{\rm Ia}/M_{\rm SF}$. Given our choice of IMF, and with $M_{\rm min} = 1$~M$_{\odot}$, $M_{\rm max}=100$~M$_{\odot}$, $M_{1}=3$~M$_{\odot}$, and $M_{2}=8$~M$_{\odot}$, this gives $\eta = 0.041$.

@@

\subsubsection{Ionizing Radiation}
\label{sec:ionizing radiation}
Radiation feedback (ionization, ionization heating, and radiation pressure) is an important source of pre-SN feedback in galaxies. HII regions carved out by stellar radiation change the ISM structure where SNe eventually explode, generally increasing their dynamical importance. However, when accounting for angular effects, radiation can also allow energy from SNe to dissipate more readily through channels carved through dense clouds. Radiation feedback effects have been included with various approximations in a wide range of simulations \citep[e.g.][]{OppenheimerDave2006, Krumholz2007, HopkinsQuataertMurray2012, Agertz2013, Renaud2013, Stinson2013, Roskar2014, Ceverino2014, FIRE, AgertzKravtsov2015, Forbes2016, Hu2016, Hu2017, FIRE2}, with limited application in full radiation hydrodynamics simulations \citep{WiseAbel2012,Wise2012a,Wise2014,Kim2013a, Kim2013b,Pawlik2013,Rosdahl2015,Aubert2015,Ocvirk2016,BaczynskiGloverKlessen2015,Pawlik2017} due to the additional computational expense of direct ray tracing. As we seek a complete accounting of stellar feedback physics, we follow \ion{H}{1} and \ion{He}{1} ionizing radiation from our stars through ray tracing methods described below. 
%mm In this particular work, we 
  We here
use two different simulations of the same galaxy to demonstrate the importance of following radiation feedback with full radiative transfer methods (see Section~\ref{sec:IC}).

@@

We adopt the \ion{H}{1} and \ion{He}{1} ionizing photon rates for each star using the OSTAR2002 \citep{Lanz2003} grid of O-type stellar models, appropriate for $M_{*} \gtrsim 15$~M$_{\odot}$ at solar metallicity\footnote{The exact stellar mass range on the OSTAR2002 grid is model dependent and a function of metallicity}. We use linear interpolation in stellar effective temperature, surface gravity, and metallicity to compute the ionizing photon fluxes and rates for each star. Stars less massive than about 15 M$_{\odot}$ are generally off of the OSTAR2002 grid, and massive stars with sub-solar metallicity generally have effective temperatures above the OSTAR2002 grid. In this case, we integrate a black body spectrum at $T_{\rm eff}$ to obtain the ionizing photon fluxes, but normalize the result to be continuous with the OSTAR2002 grid (see Appendix \ref{appendix:radiation}). \citet{Dale2012} demonstrated that uncertainties on order of a few in the ionizing radiation rates are not significant in giant molecular cloud (GMC) scale simulations.
%Generally, the ionizing fluxes from high temperature black bodies underestimate the OSTAR2002 flux, and lower temperature black bodies overestimate the flux by a factor of 2-5 and 10-1000 respectively.

Instead of assigning a fixed ionizing photon energy across all sources, we integrate over each star's blackbody curve to compute the average ionizing photon energy individually for each source (see Appendix~\ref{appendix:radiation}). The average energy for \ion{H}{1} and \ion{He}{1} ionizing photons changes significantly over the OSTAR2002 temperature range $\log(T_{4,\rm{eff}} [K]) \in \left[2.75,5.5\right]$, ranging from 15.72 eV to 20.07 eV and 26.52 eV to 31.97 eV respectively.

We also include the effects of radiation pressure on \ion{H}{1}. This has been shown to be important in suppressing the star formation rates of dwarf galaxies by influencing turbulence and the dense gas content of the ISM \citep{WiseAbel2012,Ceverino2014}. We ignore the absorption of ionizing radiation by dust and re-radiation in the infrared. This is included in other models \citep[e.g.][]{Rosdahl2015,FIRE,FIRE2} as this may increase by a factor of a few to several the effective radiation pressure \citep{ZhangDavis2017}. However, the importance of multiple scattering is still unclear. Other works have shown the effect to only increase the radiation pressure by a factor on order of unity \citep{Krumholz2012,Krumholz2013,Reissl2017}. Due to these uncertainties, and given that we focus on low metallicity dwarf galaxies with low dust content, we ignore this effect.
